---
title: "HDAnalyzeR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{HDAnalyzeR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

HDAnalyzeR is an R package developed to streamline and enhance proteomics analysis, particularly for biomarker selection from blood plasma samples. It is developed by and optimized to be used from Human Disease Blood Atlas group internally with Olink proteomics data. This vignette will guide you through the essential steps to use the package, from data loading and quality control to data cleaning, dimensionality reduction, and biomarker identification. Let's get started by loading the package!

```{r setup}
library(HDAnalyzeR)
```

This document introduces you to HDAnalyzeR's basic set of tools, and shows you how to analyze and identify biomarkers in a dataset of cancer blood plasma samples.

## Loading the Data

First, we load the package's `example_data` and `example_metadata`.

```{r}
head(example_data)
```

```{r}
head(example_metadata)
```

> 📓 In real-world scenarios, you would load your own data and metadata files instead of using the example dataset. In order to run the package without issues, make sure that your data include the following columns: DAid, Assay and, NPX, while your metadata include the following columns: DAid and Disease. Also, if you have a Sex column in your metadata, the data should be encoded as M and F.

## Quality Control (QC)

### Data QC

`qc_summary_data()` provides a comprehensive summary of the input dataset. It will check the column types, calculate the percentage of NAs in each column and row, perform normality tests for all the different Assays, calculate protein-protein correlations, and create a heatmap of these correlations. Users can also specify the threshold for reporting protein-protein correlations.

```{r}
qc_data <- qc_summary_data(example_data, wide = FALSE, threshold = 0.7)
qc_data$heatmap
```

### Metadata QC

`qc_summary_metadata()` summarizes quality control results of the metadata dataframe. It checks the column types, calculates the percentage of NAs in each column and row exactly as `qc_summary_data()`, and creates summary visualizations for key metadata variables such as Sex, Age, and BMI.

```{r}
qc_summary_metadata(example_metadata, disease_palette = "cancers12")
```

## Data Cleaning

### Data Cleaning

As we saw from the QC results, the data contains NAs and other issues that need to be addressed. `clean_data()` preprocesses the dataset by filtering out rows based on specified criteria. In this case we will keep only the data which Assay_Warning is "PASS" and only DAid, Assay and NPX columns.

```{r}
clean_data <- clean_data(example_data, 
                         keep_cols = c("DAid", "Assay", "NPX"),
                         filter_assay_warning = TRUE)
head(clean_data)
```

### Metadata Cleaning

In our case, `clean_metadata()` preprocesses the metadata just by keeping only the specified columns.

```{r}
clean_metadata <- clean_metadata(example_metadata, 
                                 keep_cols = c("DAid", "Disease", "Sex", "Age"))
head(clean_metadata)
```

### Data Transformation

Once the data is cleaned, we recommend transforming it into a tidy format (wide format) if it's not already in that form. `generate_df()` will create for us both the wide Olink dataset as well as the joined with metadata dataset.

```{r}
dfs <- generate_df(clean_data, 
                   clean_metadata, 
                   metadata_cols = c("DAid", "Disease", "Sex", "Age"), 
                   save = FALSE)

wide_data <- dfs$wide_data
head(wide_data)

join_data <- dfs$join_data
head(join_data)
```

> 📓 While HDAnalyzeR can work with long format data, most functions will transform it into a wide format, which can slightly slow down the pipeline. Thus, starting with tidy data is advisable.

## Imputation and Dimensionality Reduction

Next, we will impute missing values using K-nearest neighbors (KNN) with 3 neighbors via `impute_knn()`.

```{r}
imputed_data <- impute_knn(wide_data, 
                           k = 3,
                           exclude_cols = c("DAid"),
                           show_na_percentage = FALSE)
head(imputed_data)
```

After imputation, we will run Principal Component Analysis (PCA) via `do_pca()` and Uniform Manifold Approximation and Projection (UMAP) via `do_umap()` to check for outliers, batch effects, and other potential issues.

```{r}
do_pca(imputed_data, 
       clean_metadata,
       color = "Sex",
       palette = "sex_hpa",
       impute = FALSE,
       pcs = 6)
```

```{r}
do_umap(imputed_data, 
        clean_metadata,
        color = "Disease",
        palette = "cancers12",
        impute = FALSE)
```

We will perform another QC check to ensure that everything is as expected after cleaning and imputing the data.

```{r}
qc_data <- qc_summary_data(imputed_data, wide = TRUE, threshold = 0.7)
qc_data$heatmap
```

```{r}
qc_summary_metadata(clean_metadata, disease_palette = "cancers12")
```

## Biomarker Identification

### Differential Expression Analysis

We will run a differential expression analysis to identify potential biomarkers. We will use `do_limma()` so that we will be able to correct also for Age. This method will help us pinpoint proteins that are significantly different between conditions. We will present only the results for AML (Acute Myeloid Leukemia), BRC (Breast Cancer), and PRC (Prostate Cancer).

```{r}
de_res_aml <- do_limma(imputed_data, 
                       clean_metadata,
                       case = "AML",
                       control = c("BRC", "PRC"),
                       correct = c("Sex", "Age"),
                       correct_type = c("factor", "numeric"),
                       only_female = "BRC",
                       only_male = "PRC")

de_res_aml$de_results

de_res_aml$volcano_plot
```

```{r}
de_res_brc <- do_limma(imputed_data, 
                       clean_metadata,
                       case = "BRC",
                       control = c("AML", "PRC"),
                       correct = c("Sex", "Age"),
                       correct_type = c("factor", "numeric"),
                       only_female = "BRC",
                       only_male = "PRC")

de_res_brc$de_results

de_res_brc$volcano_plot
```

```{r}
de_res_prc <- do_limma(imputed_data, 
                       clean_metadata,
                       case = "PRC",
                       control = c("BRC", "AML"),
                       correct = c("Sex", "Age"),
                       correct_type = c("factor", "numeric"),
                       only_female = "BRC",
                       only_male = "PRC")

de_res_prc$de_results

de_res_prc$volcano_plot
```

We can also summarize the results via `plot_de_summary()`.  In order to use this function we need to store the results in a list.

```{r}
de_res <- list("AML" = de_res_aml, 
               "BRC" = de_res_brc, 
               "PRC" = de_res_prc)

plot_de_summary(de_res, disease_palette = "cancers12")
```

### Lasso Machine Learning Classification Model

In addition to differential expression analysis, we will use a Lasso machine learning classification model to identify significant features. This model will help us understand which proteins are most predictive of the conditions being studied. Once again, we will present only the results for AML, BRC and, PRC.

```{r}
lasso_res_aml <- do_elnet(imputed_data, 
                          clean_metadata, 
                          case = "AML",
                          control = c("BRC", "PRC"),
                          only_female = "BRC",
                          only_male = "PRC",
                          exclude_cols = c("Sex", "Age"),
                          type = "lasso",
                          palette = "cancers12",
                          subtitle = c("accuracy", 
                                       "sensitivity", 
                                       "specificity", 
                                       "auc", 
                                       "features",
                                       "top-features"),
                          nfeatures = 12,
                          points = FALSE)

lasso_res_aml$hypopt_res$hypopt_vis

lasso_res_aml$testfit_res$metrics

lasso_res_aml$var_imp_res$features

lasso_res_aml$var_imp_res$var_imp_plot

lasso_res_aml$boxplot_res
```

```{r}
lasso_res_brc <- do_elnet(imputed_data, 
                          clean_metadata, 
                          case = "BRC",
                          control = c("AML", "PRC"),
                          only_female = "BRC",
                          only_male = "PRC",
                          exclude_cols = c("Sex", "Age"),
                          type = "lasso",
                          palette = "cancers12",
                          subtitle = c("accuracy", 
                                       "sensitivity", 
                                       "specificity", 
                                       "auc", 
                                       "features",
                                       "top-features"),
                          nfeatures = 12,
                          points = FALSE)

lasso_res_brc$hypopt_res$hypopt_vis

lasso_res_brc$testfit_res$metrics

lasso_res_brc$var_imp_res$features

lasso_res_brc$var_imp_res$var_imp_plot

lasso_res_brc$boxplot_res
```

```{r}
lasso_res_prc <- do_elnet(imputed_data, 
                          clean_metadata, 
                          case = "PRC",
                          control = c("BRC", "AML"),
                          only_female = "BRC",
                          only_male = "PRC",
                          exclude_cols = c("Sex", "Age"),
                          type = "lasso",
                          palette = "cancers12",
                          subtitle = c("accuracy", 
                                       "sensitivity", 
                                       "specificity", 
                                       "auc", 
                                       "features",
                                       "top-features"),
                          nfeatures = 12,
                          points = FALSE)

lasso_res_prc$hypopt_res$hypopt_vis

lasso_res_prc$testfit_res$metrics

lasso_res_prc$var_imp_res$features

lasso_res_prc$var_imp_res$var_imp_plot

lasso_res_prc$boxplot_res
```

We can get a visual summary of the results via `plot_features_summary()` too. In order to use this function we need to store the results in a list.

```{r}
lasso_res <- list("AML" = lasso_res_aml, 
                  "BRC" = lasso_res_brc, 
                  "PRC" = lasso_res_prc)

plot_features_summary(lasso_res, disease_palette = "cancers12")
```

## One step further

The final step involves performing an Over-Representation Analysis for the up-regulated proteins from differential expression, that are also identified as features by the ML model. In this example, we will use the Gene Ontology (GO) database and show the results only for AML.

> 📓 Remember that these data are a dummy-dataset with fake data.

```{r}
# Extract the proteins identified by both DE and Lasso
de_proteins <- de_res_aml$de_results |> 
  dplyr::filter(sig == "significant up") |> 
  dplyr::pull(Assay)

lasso_proteins <- lasso_res_aml$var_imp_res$features |> 
  dplyr::filter(Scaled_Importance > 0) |> 
  dplyr::pull(Variable)

intersect_proteins <- intersect(de_proteins, lasso_proteins)

# Perform ORA with GO database and visualize results
enrichment <- do_ora(intersect_proteins, database = "GO")
plot_ora(enrichment, intersect_proteins, ncateg = 5)
```
