---
title: "Post Analysis: Pathway Enrichment & Automated Literature Search"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{post_analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette will guide you through the post analysis of the results obtained from the HDAnalyzeR pipeline. The post analysis consists of two possible steps: pathway enrichment analysis and automated literature search. The pathway enrichment analysis is performed using the Gene Ontology and Reactome databases from `clusterProfiler` and `ReactomePA` packages respectively. The automated literature search is performed using the the PubMed database.

Let's start by loading the packages, widen the example dataset, load the metadata, and run some differential expression analysis to get the results for the post analysis. For the Over Representation Analysis we could also use the features list from the classification models or even run both and get the intersect as it is done in the Get Started guide.

```{r setup, include = FALSE}
library(HDAnalyzeR)
library(dplyr)
library(utils)

wide_data <- widen_data(example_data)
metadata <- example_metadata
de_res <- do_limma(wide_data, metadata, case = "AML", control = c("CLL", "MYEL", "GLIOM"))
```

First, we will perform an Over Representation Analysis (ORA) using the Gene Ontology database. We will use the `do_ora()` function. The function requires a list of proteins. In this example we will extract the top-20 up-regulated proteins.

```{r}
proteins <- de_res$de_results |> filter(logFC > 0) |> pull(Assay) |> head(20)

ora_res <- do_ora(proteins, "GO")
plot_ora(ora_res, proteins)
```
> 📓 When working with a real Olink dataset, it is recommended to use a background list of proteins that are present in the Olink panel to minimize bias. This can be done by using the `background` parameter in the `do_ora()` function.

Let's change the database and the p-value threshold.

```{r, include = FALSE}
ora_res <- do_ora(proteins, "Reactome", pval = 0.2)
plot_res <- plot_ora(ora_res, proteins, pval = 0.2, fontsize = 6)
```
```{r}
plot_res$dotplot
plot_res$barplot
plot_res$cnetplot
```

We can also run a Gene Set Enrichment Analysis (GSEA) using the `do_gsea()` function. In this case, the function requires differential expression results. 

```{r, include = FALSE}
gsea_res <- do_gsea(de_res$de_results, "GO", pval_lim = 0.7)
plot_res <- plot_gsea(gsea_res, de_res$de_results, pval_lim = 0.7, fontsize = 7)
```
```{r}
plot_res$dotplot
plot_res$cnetplot
plot_res$ridgeplot
plot_res$gseaplot
```

Finally, let's perform an automated literature search using the `literature_search()`. The function requires a list with disease names as names and proteins as values. We will create the list, run the search and preview the results.

```{r}
biomarkers <- list("acute myeloid leukemia" = c("FLT3", "EPO"),
                   "chronic lymphocytic leukemia" = c("PARP1", "FCER2"))

lit_res <- literature_search(biomarkers, max_articles = 5)

lit_res$`acute myeloid leukemia`$FLT3$title
```

> 📓 Remember that these data are a dummy-dataset with fake data and the results in this guide should not be interpreted as real results. The purpose of this vignette is to show you how to use the package and its functions.
